Human capital mediates natural selection in contemporary humans

### 2022-05-04 response-to-reviewers-BE.Rmd

> abstract: "Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against)." This is hard to understand. How about: "Consistently over time, polygenic scores predicting higher earnings, education, and health also predict lower fertility."

Done.

> p\. 2: "natural selection may substantially increase the correlation between genetic scores and income, i.e. the "genetic lottery" (Harden, 2021)." Although the meaning eventually becomes clear, this sentence may be very cryptic.

We have made it more explicit: "natural selection substantially increases the correlation between polygenic scores and income, increasing genetic differences between different social groups, and thus making the "genetic lottery" (Harden 2021) more unfair."

@@

The functional form is a standard one in economics. It has the specific property that relative risk aversion is constant. That is, if you have CRRA utility and £100, and you are indifferent to a 50-50 bet between winning £30 and losing £20, then if you had £1000 you would be indifferent between a 50-50 to win £300 or lose £200.

The literature on estimating utility functions is vast (see e.g. Chuang and Schechter 2015). One issue is that of context. CRRA may make sense for risky financial decisions. Here, as Reviewer 2 points out, the "curvature" of the utility function is used not to explain attitudes to risk, but to trade off income against children at different income levels. This author wishes there was a single utility function that worked to explain all human behaviour across contexts... in reality, economists tend to use "what works". We use the CRRA form because it is tractable; we expect that using other forms, intermediate levels of risk aversion would still generate our empirical results, but the maths would be uglier. For a useful discussion of microeconomic theories of fertility, including functional forms, see Jones et al. (2008) and more recently Doepke et al. (2022) which uses a very similar form to ours to capture birth timing.

> p\. 30: I feel that more thought might be given to the choice of covariates. For example, if fluid IQ is on the causal path between the PGS and EA, then controlling it will bias downward the estimate of the the PGS effect on EA. But since this mediation analysis seems tentative and not crucial to the paper, perhaps some looseness here can be tolerated.

Yes. Conversely, fluid IQ might affect fertility choices via other pathways than human capital in the labour market, and not controlling for it might leave these to be picked up by EA. We prefer to be "conservative" in controlling for it, in the sense that we'd rather bias the estimate indirect effect of EA towards zero than away from zero. We now mention this issue in the appendix: "Note that controlling for fluid IQ is conservative, in the sense that IQ might also measure human capital in the labour market; nevertheless, we control for it because fluid IQ might affect fertility decisions via other channels than human capital."

@@

> p\. 34: For various reasons I think this section (and the bit in the main text that it supports) ought to be eliminated. The point is to show that natural selection increases the correlation between the EA PGS and income from one generation to the next, correcting for the noise in the PGS. I think a qualitative statement about this point is adequate; the more quantitative attempt by the authors has some problems.
>
> (1) "PSEA" is not defined, although I'm guessing this means "polygenic score for educational attainment."
>
> (2) It happens that there is a more recent meta-analysis of relevant twin studies (Silventoinen et al., 2020).
>
> (3) There appears to be substantial passive gene-environment correlation ("genetic nurture," some now call it), as a result of highly educated parents providing both favorable genes and a social status promoting educational attainment (e.g., Kong et al., 2018). I am sure this complicates the analysis in some way. The authors might respond that they deal with this by using within-family estimates, but see below.
>
> (4) The proof of Equation 5 uses the "standard errors-in-variables formula" plim \\hat\\beta = \\lambda\\beta. While this relationship seems plausible, it may not be standard to behavioral geneticists.
>
> (5) When there is genetic nurture, what is the right PGS effect to use in order to address the question here? The effect in the population, or the effect within families? If it is the latter, a problem is that the estimate of a 142.6% change in R\^2 between generations seems far too high and indicative of low statistical precision.
>
> (6) The assumption that eta is independent of p_star seems very strong. It includes, for example, the assumption that the PGS over rare variants is uncorrelated with the PGS over common variants. I think this is unlikely to be true, as a result of assortative mating (Yengo et al., 2018).

We accept point (6) in particular. We had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be. The within-family estimate should remove many of the worries about exogeneity of p\*, but it is estimated with a smaller sample. We agree that the estimate seems high, and indeed, the bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards. We've removed the discussion.

@@

We don't want to go too far, by pretending that the theory came to us fully-formed in our ivory tower, and then we tested it. In fact, the motivating results led us to develop the theory, first informally and then more precisely; only the results in section 4 should be considered a clean test.

> In my opinion, the paper's most important finding is that human capital mediates natural selection. I suspect the authors agree with me on this point, since the title of their manuscript says just that. Still, when reading the paper, one does not get to the topic of human capital and natural selection until p. 9 (and the paper has only 15 pages)! The lengthy discussion of the stratified analysis (Figures 3-5) is interesting, and the authors have edited the text (vs. in the earlier version) to explain that these results are related to predictions of their model, but that discussion is too long and distracts from the main point of the paper. Plus, the stratified analyses are\
> limited and only so much can be learned from them. As the authors acknowledge, "these categories are not necessarily exogenous to the polygenic scores. For example, both in the data and in our theoretical model, age at first live birth is a choice variable, which is endogenous to human-capital-related polygenic." Thus, I recommend that that discussion be shortened and the figures consolidated (the authors should find a way to show the must important results in Figures 3-5 on \~1 page.

We think these analyses are important. First, they motivate our theory, in which education, income, AFLB etc. are indeed endogenous choice variables which are co-determined with fertility. If these mediators were *not* endogenous, we'd need a different theory.

We also think the results matter in their own right. Any theory of contemporary natural selection needs to explain why correlations with fertility are lower at higher levels of income, education, etc. Non-causal explanations are fine, but theories which don't generate these correlations must be missing something. Note that Fisher, cited by Reviewer 1, used observed correlations quite like ours to differentiate between theories of the fertility-income relationship. Generally, significant correlations have often led to important theoretical literatures, even when causality is initially uncertain (e.g. education/wages in labour economics, or democracy/GDP levels in political economy).

@@

As described above, we have shortened Section 2 to focus on the key results of variation by income, education, number of partners, partner presence and AFLB. These empirics now take up just 39 lines of text, plus Figures 3-5. We no longer split respondents by sex in Figure 4. This saves two figures, and should reduce space in the final version.

> 2\. The authors test their human-capital-mediation theory in two ways in Section 4, but I think their analysis can and should be improved, since this section contains the core empirical results supporting the authors' main claim about human capital mediation. First, the authors show that a PGS's correlation with two measures of human capital negatively correlates with selection effects for the PGS (Figure 6); second, they run a formal mediation analysis.
>
> \> Can the authors introduce a basic econometric framework and assess whether their results in Figure 6 are consistent with those from the mediation analysis (i.e., could the amount of mediation reported in Table 5 for each phenotype explain the results shown in Figure 6)?

The result in Figure 6 is that the size of a score's selection effect correlates with the score's correlation with education and earnings. Intuitively, the fact that part of the selection effect comes from the mediation indeed helps to explains table 6. That is, if

*PGS effect on fertility = PGS correlation with education* $\times$ *education effect on fertility + direct effect of PGS on fertility*

@@

then indeed PGS effect on fertility, and PGS correlation with education, will be correlated. In that sense, the mediation analysis is just a more precise version of Figure 6 (more precise because it also estimates the education effect on fertility). We're not sure if there is an obvious way to work out "how much" of the Figure 6 correlation is explained by mediation. It may be a bit like asking "how much of a bivariate correlation is explained by the same correlation but with controls": these are really two alternate models of the same process, one less controlled than than the other. We are probably hitting the limits of our econometric expertise, so we are open to suggestions.

> \> In the mediation analysis, why do the authors only use education as the mediator? The mediation framework they use can easily accommodate many mediators (see Section 6 of the Supplementary Information of the "EA2" paper by Okbay et al., Nature Genetics 2016). I would like to see a more full-fledged mediation analysis that also includes other proxies for human capital as mediators, including earnings in a respondent's first job as well as household income (though the later is not an ideal variable for that, especially for females).

We would like to have more mediators, but appropriate variables are hard to find. Earnings in first job reduces our N by a factor of 10, and this makes results too noisy to be informative. As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator. We welcome any other ideas. See also our response to Reviewer 1's comment on fluid IQ.

@@

> OTHER COMMENTS AND SUGGESTIONS
>
> 8\. Most figures are hard to read. For example, Figures 3-5 are hard to read. In Figure 3a, the dot color indicate education and the dot border color indicates significance; that's very difficult to read quickly (for instance, a blue dot with a yellow border looks green). Instead, the authors could use different symbols (crosses, squares) for education and colors for significance, or vice-versa.

We have changed our figures to use both shape and colour for categories. We hope this is an improvement. We've also worked on Figure 1.

> 9\. The authors should provide more info to help readers understand the details of the analyses whose results are shown in Appendix Tables 3-4.

We now say: 'We report the number of scores showing significant changes over time (i.e. a significant interaction between polygenic score and the "late born" dummy): either a significant change in sign, a significant increase in effect size, or a significant decrease in size.'

### 2022-05-04 why-natural-selection.Rmd
typos and fixes for edits

Figure \@ref(fig:plot-sexes) plots selection effects by sex. Differences are
particularly large for educational attainment, height and MDD. Several polygenic
scores for mental illness and personality traits are more selected for (or less
against) among women, including major depressive disorder (MDD), schizophrenia
and neuroticism, while extraversion is more selected for among men. Scores for
waist circumference and waist-hip ratio are less selected for among women. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

We report results for 3 alternative weighting schemes: weighting by
age/qualification; geographical (weighting by Middle Super Output Area); and for
women only, age, qualification and age at first live birth. Population data for
weighting is taken from the 2011 UK Census and the 2006 General Household Survey
(GHS). Weighting for Age/Qualification and Age/Qualification/AFLB weights was
done using marginal totals from a linear model, using the `calibrate()` function
in the R "survey" package [@lumley2020survey]. Geographical weighting was done
with iterative post-stratification using the `rake()` function, on Census Middle
Layer Super Output Areas, sex and presence/absence of a partner.

@@

Table \@ref(tab:tbl-change-by-period) 
summarizes the results. We report the number of scores showing significant
changes over time (i.e. a significant interaction between polygenic score and
the "late born" dummy): either a significant change in sign, a significant
increase in size, or a significant decrease in size. There is little evidence
for changes in selection effects within the parents' generation, with just one
score showing a significant decrease in size. In the respondents' generation,
effect sizes were significantly larger in absolute size among the later-born for
eight polygenic scores: ADHD, age at menopause, cognitive
ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
Disorder. These changes are inconsistent with the intergenerational change,
where estimated effect sizes were larger among the earlier, parents' generation.

@@

Figure \@ref(fig:plot-age-birth-parents) shows the regressions controlling for
either parent's age at their birth. Effect sizes are very similar, whether
controlling for father's or mother's age. As in the respondents' generation,
effect sizes are negatively correlated with the effect sizes from bivariate
regressions without the age at birth control (father's age at birth: $\rho$ 
`r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).

>>>>

Figure \@ref(fig:plot-sexes) plots selection effects by sex. Differences are
particularly large for educational attainment, height, ADHD and MDD. Several
polygenic scores for mental illness and personality traits are more selected for
(or less against) among women, including major depressive disorder (MDD),
schizophrenia and neuroticism, while extraversion is more selected for among
men.

@@

We compare results for our main weights to 3 alternative weighting schemes:
weighting by age/qualification; geographical (weighting by Middle Super Output
Area); and for women only, age, qualification and age at first live birth.
Population data for weighting is taken from the 2011 UK Census and the 2006
General Household Survey (GHS). Weighting for Age/Qualification and
Age/Qualification/AFLB weights was done using marginal totals from a linear
model, using the `calibrate()` function in the R "survey" package
[@lumley2020survey]. Geographical weighting was done with iterative
post-stratification using the `rake()` function, on Census Middle Layer Super
Output Areas, sex and presence/absence of a partner.

@@

Table \@ref(tab:tbl-change-by-period) summarizes the results. We report the
number of scores showing significant changes over time (i.e. a significant
interaction between polygenic score and the "late born" dummy): either a
significant change in sign, a significant increase in effect size, or a
significant decrease in size. There is little evidence for changes in selection
effects within the parents' generation, with just one score showing a
significant decrease in size. In the respondents' generation, effect sizes were
significantly larger in absolute size among the later-born for eight polygenic
scores: ADHD, age at menopause, cognitive ability, Coronary Artery Disease, EA2,
EA3, extraversion and Major Depressive Disorder. These changes are inconsistent
with the intergenerational change, where estimated effect sizes were larger
among the earlier, parents' generation.

@@

Figure \@ref(fig:plot-age-birth-parents) shows the regressions controlling for
either parent's age at their birth. Effect sizes are very similar, whether
controlling for father's or mother's age. As in the respondents' generation,
effect sizes are negatively correlated with the effect sizes from bivariate
regressions without the control for age at birth (father's age at birth: $\rho$ 
`r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).
